Modeling Second-Language Learning from a Psychological Perspective

### 2018-03-28 doc/naaclhlt2018-latex/paper.tex
proofreading

\author{Alexander S. Rich\qquad Pamela Osborn Popp\qquad David J. Halpern\\
  \textbf{Anselm Rothe\qquad Todd M. Gureckis} \\
  Department of Psychology, New York University \\
  {\tt \{asr443,pamop,david.halpern,anselm,todd.gureckis\}@nyu.edu} \\}

@@

Educational software that aims to teach people new skills, languages, and
academic subjects have become increasingly popular.  The wide-spread
deployment of these tools has created interesting opportunities to study
the process of learning in extremely large samples of learners in naturalistic
situations. The Duolingo shared task on Second Lanugage Acquisition Modeling (SLAM)
was a competitive modeling challenge run in early 2018~\cite{slam18}.
The challenge, organized by Duolingo\footnote{http://duolingo.com}, a popular second
language learning app, was
to use log data from thousand of users completing millions of exercises 
predict patterns of future translation mistakes in held-out data.  The data set was
divided into three sets covering Spanish speakers learning English ({\tt en\_es}),
English speakers learning Spanish ({\tt es\_en}), and English speakers learning
French ({\tt fr\_en}).
This paper reports the approach used by our team which ended in third place 
for the {\tt en\_es} data set, second place for {\tt es\_en}, and third place
for {\tt fr\_en}.

Our lab's interest in this competition comes from our emphasis on studying human learning
and memory from a psychological perspective.  While the Duolingo SLAM dataset 
heavily emphasizes language learning (e.g., syntax, grammar, semantics), it also touches on
basic learning and memory phenomena (e.g., retrieval of the translated
word from memory).  The study of the psychological processes supporting second-language 
acquisition has long been a topic of interest in the learning and memory literature~\cite[e.g.,]{Atkinson:1972rm,Atkinson:1972rz,Pavlik:2008rm}.
In fact, much of the classic work on psychology of memory involved aspects of verbal
language learning, which helped to identify a number of important and robust phenomena
in memory including the retention curve and the advantage for spaced
over massed practice~\cite{Ruth:1928aa,Rubin:1996aa,Cepeda:2006aa}.

@@

Along with the tokens themselves we encoded \hl{each} instance word's part of speech,
morphological features, and dependency edge label. (We noticed that some words
in the original dataset were paired with the wrong morphological features,
particularly near where punctuation had been removed from the sentence. To fix
this, we re-processed the data using Google SyntaxNet\footnote{https://github.com/ljm625/syntaxnet-rest-api}.)

We also encoded word length and several word characteristics gleaned from
external data sources. The word frequency effect suggests that uncommon words are harder to process than common words; readers will look longer at low-frequency words and perform worse in word-identification tasks for these than for high-frequency words \cite{rayner1998eye}. We therefore included a feature that encoded the frequency of each word in the language being acquired, calculated from \citet{robert_speer_2017_998161}. A number of studies have also shown that age-of-acquisition is another strong predictor of word processing and lexical retrieval difficulty that is somewhat independent of word frequency \cite{brysbaert2011effects, ferrand2011comparing}. We therefore included mean age-of-acquisition (in English) as a feature, derived from published age-of-acquisition norms for 30,000 English words from \cite{Kuperman2012} which covered many of the words present in the dataset. Additionally, cognates, or words sharing a common linguistic derivation, are easier to learn than words with dissimilar translations \cite{de2000hard}. As an approximate measure of linguistic similarity, we used the Levenshtein edit distance between the word tokens and their translations scaled by the length of the longer word. We found translations using Google Translate\footnote{https://cloud.google.com/translate/} and calculated the Levenshtein distance to reflect the letter-by-letter similarity of the word and its translation \cite{hyyro2001explaining}. 
\subsubsection{User features}

@@

To estimate users' diligence we speculated that a very diligent user might be using the app regularly at the same time of day, perhaps following a study schedule, compared to a less diligent user whose times might vary a lot. The data set did not provide a variable with the time of day, which would have been an interesting feature on its own. Instead, we were able to extract for each exercise the time of day relative to the first time a user had used the app, ranging from 0 to 1 (with 0 indicating the same time, 0.25 indicating a relative shift by 6 hours, etc.).  We then discretized this variable into 20-minute bins and computed the entropy of the empirical frequency distribution over these bins. A lower entropy score indicated less variability in the times of day a user started their exercises.

@@

Interestingly, we found that for both user-level and word-level features, the
bulk of the model's predictive power could be achieved using ID's alone,
represented as high-cardinality categorical features. Removing other word features,
such as morphological features and part of speech, created only a small
degradation of performance. In the case of users, removing features such as
entropy and average exercise burst length led to a tiny increase of performance.
In the case of both users and words, though, we find that in the absence of ID
features the other features are helpful and leads to better performance than
removing all features. We also found that removing all information about
neighboring words and the dependency-parse root word degraded performance. This
confirms that word context matters, and likely points to errors in word order,
subject--verb matching and other grammatical rules.

@@

There are multiple reasons for the limited applicability of psychology to this
competition. First, computational psychological models, while useful, are not
well-suited to generating highly accurate predictions from large data sets.
Because they are designed not for prediction but for explanation, they tend to
use a small number of input variables and to allow those variables to interact
in limited ways. In contrast, gradient boosted trees, as well as other
cutting-edge techniques like deep learning. can extract high-level
interactions among hundreds of features. Though they are highly opaque, require
a lot of data, and are not amenable to explanation, these models excel at prediction.

>>>>

\author{Alexander S. Rich\qquad Pamela J. Osborn Popp\qquad David J. Halpern\\
  \textbf{Anselm Rothe\qquad Todd M. Gureckis} \\
  Department of Psychology, New York University \\
  {\tt \{asr443,pamop,david.halpern,anselm,todd.gureckis\}@nyu.edu} \\}

@@

Educational software that aims to teach people new skills, languages, and
academic subjects have become increasingly popular.  The wide-spread
deployment of these tools has created interesting opportunities to study
the process of learning in extremely large samples of learners in naturalistic
situations. The Duolingo shared task on Second Lanugage Acquisition Modeling (SLAM)
was a competitive modeling challenge run in early 2018~\cite{slam18}.
The challenge, organized by Duolingo\footnote{http://duolingo.com}, a popular second
language learning app, was
to use log data from thousands of users completing millions of exercises to 
predict patterns of future translation mistakes in held-out data.  The data set was
divided into three sets covering Spanish speakers learning English ({\tt en\_es}),
English speakers learning Spanish ({\tt es\_en}), and English speakers learning
French ({\tt fr\_en}).
This paper reports the approach used by our team which ended in third place 
for the {\tt en\_es} data set, second place for {\tt es\_en}, and third place
for {\tt fr\_en}.

Our lab's interest in this competition comes from our emphasis on studying human learning
and memory from a psychological perspective.  While the Duolingo SLAM dataset 
heavily emphasizes language learning (e.g., syntax, grammar, semantics), it also touches on
basic learning and memory phenomena (e.g., retrieval of the translated
word from memory).  The study of the psychological processes supporting second-language 
acquisition has long been a topic of interest in the learning and memory literature~\cite[e.g.,]{Atkinson:1972rm,Atkinson:1972rz,Pavlik:2008rm}.
In fact, much of the classic work on the psychology of memory involved aspects of verbal
language learning, which helped to identify a number of important and robust phenomena
in memory including the retention curve and the advantage for spaced
over massed practice~\cite{Ruth:1928aa,Rubin:1996aa,Cepeda:2006aa}.

@@

Along with the tokens themselves we encoded each instance word's part of speech,
morphological features, and dependency edge label. (We noticed that some words
in the original dataset were paired with the wrong morphological features,
particularly near where punctuation had been removed from the sentence. To fix
this, we re-processed the data using Google SyntaxNet\footnote{https://github.com/ljm625/syntaxnet-rest-api}.)

We also encoded word length and several word characteristics gleaned from
external data sources. The word frequency effect suggests that uncommon words are harder to process than common words; readers will look longer at low-frequency words and perform worse in word-identification tasks for these than for high-frequency words \cite{rayner1998eye}. We therefore included a feature that encoded the frequency of each word in the language being acquired, calculated from \citet{robert_speer_2017_998161}. A number of studies have also shown that age-of-acquisition is another strong predictor of word processing and lexical retrieval difficulty that is somewhat independent of word frequency \cite{brysbaert2011effects, ferrand2011comparing}. We therefore included mean age-of-acquisition (in English) as a feature, derived from published age-of-acquisition norms for 30,000 English words from \cite{Kuperman2012} which covered many of the words present in the dataset. Additionally, cognates, or words sharing a common linguistic derivation, are easier to learn than words with dissimilar translations \cite{de2000hard}. As an approximate measure of linguistic similarity, we used the Levenshtein edit distance between the word tokens and their translations scaled by the length of the longer word. We found translations using Google Translate\footnote{https://cloud.google.com/translate/} and calculated the Levenshtein distance to reflect the letter-by-letter similarity of the word and its translation \cite{hyyro2001explaining}.

\subsubsection{User features}

@@

To estimate users' diligence\hl{consistency?} we speculated that a very diligent user might be using the app regularly at the same time of day, perhaps following a study schedule, compared to a less diligent user whose schedule might vary more. The data set did not provide a variable with the time of day, which would have been an interesting feature on its own. Instead, we were able to extract for each exercise the time of day relative to the first time a user had used the app, ranging from 0 to 1 (with 0 indicating the same time, 0.25 indicating a relative shift by 6 hours, etc.).  We then discretized this variable into 20-minute bins and computed the entropy of the empirical frequency distribution over these bins. A lower entropy score indicated less variability in the times of day a user started their exercises.

@@

Interestingly, we found that for both user-level and word-level features, the
bulk of the model's predictive power could be achieved using ID's alone,
represented as high-cardinality categorical features. Removing other word features,
such as morphological features and part of speech, created only a small
degradation of performance. In the case of users, removing features such as
entropy and average exercise burst length led to a tiny increase of performance.
In the case of both users and words, though, we find that in the absence of ID
features the other features are helpful and leads to better performance than
removing all features. We also found that removing all information about
neighboring words and the dependency-parse root word degraded performance. This
confirms that word context matters, and suggests that users commonly make errors in word order,
subject--verb matching and other grammatical rules.

@@

There are multiple reasons for the limited applicability of psychology to this
competition. First, computational psychological models, while useful, are not
well-suited to generating highly accurate predictions from large data sets.
Because they are designed not for prediction but for explanation, they tend to
use a small number of input variables and allow those variables to interact
in limited ways. In contrast, gradient boosted trees, as well as other
cutting-edge techniques like deep learning. can extract high-level
interactions among hundreds of features. Though they are highly opaque, require
a lot of data, and are not amenable to explanation, these models excel at prediction.
